It's all about Meaning!\\{\normalsize A Theory for Data Migration when Deploying Changes

### 2022-12-18 2022 Migration/articleMigration.tex
fixed mistake in definition 7 (inst)

\begin{abstract}
   To support incremental software development with software generators bears the potential of decreasing the time to deploy changes.
   However, an increment that changes the database schema causes a nontrivial data migration problem.
   The problem is how to preserve the semantics of the data as much as possible and satisfy elementary business requirements at the same time.

   The Ampersand project has provided the theory and tools to generate information systems from an algebraic specification.
   This paper develops a theory for making changes to an already deployed information system that was generated by Ampersand.
   We propose a migration method with zero down-time, and allows the business to undo the change after verification.
   It also leaves room for the business to finish ongoing work after the moment the change is deployed.
   We aim to automate the change including the data migration,
   to prevent mistakes and enable more frequent changes.
\end{abstract}

@@

   We believe in generating information systems%
\footnote{In the sequel, the word ``system'' refers to the phrase ``information system''. This simplifies the language a little. }
   to prevent human programming errors.
   For, a mistake not made can never propagate into production.
   We also believe that information systems should be released into production in small increments,
   to obtain shorter release cycles, more frequent releases, and thus a more agile software development process%
   \footnote{These effects can be measured in terms of DevOps metrics such as
   deployment frequency,
   reliability of deployments, and
   change failure rate~\cite{DevOps2021}.}.
   By installing upgrades automatically, frequently, in small increments, and without down-time,
   the users experience a system that evolves organically, without even noticing the updates.
   Thus, we believe our theory will help to automate data migrations correctly and reliably.
   The benefits work in two directions.
   Automation data migrations saves effort and errors,
   contributing to more frequent and smaller releases.
   Smaller and faster releases also mean a smaller difference between the old system and the new system.
   The smaller that difference, the simpler the migration can be and the more of it can be automated.

   In each increment, we distinguish an ``old system'' and a ``new system''.
   A difference of schemas complicates the migration of data between the old system and the new system.
   The purpose of incremental evolution poses specific requirements to data migration,
   which sets it apart from data migrations for other purposes.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers may exclude new functionality to avoid introducing new errors in an otherwise error-prone migration process.
   In our work, however, introducing new functionality and removing obsolete functionality is the very purpose of an increment.
   Data migration literature sometimes uses a different notion of increment.
   For example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014} define an increment as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   In our work, however, we want to get rid of the old system and not retain it.
   So, we need to look at a different approach.
   Examples like these have convinced us to define a new method for data migration that is specifically meant for systems in production that are
   continually evolving in small increments.

   The contribution of this paper is to derive a migration schema from two artifacts: an existing system
   (which has its own schema and dataset) and a new schema (which specifies the new system),
   which allows us to facilitate the migration.
   The purpose of our theory is to develop tools.
   These tools must satisfy such practical requirements as
   zero down-time updates and ways to cope with data pollution.
   In practice, more frequent updates require that updates come with zero down-time.
   For our theory, zero down-time means that all human activity for the data migration can be done without taking the system out of production.
   We consider any delay caused by automatically switching from the old system to the new one negligible,
   assuming a modern deployment platform that does this with zero down-time.

@@

\section{Analysis}
   The purpose of an information system is to make data meaningful to its users.
   Users have their own tasks and responsibilities
   and may work from different locations and on different moments.
   This collective use serves a purpose which we loosely call ``the business''.
   As a consequence, the data in a system changes continually.
   To preserve meaning, users are trying to maintain semantic constraints on the data,
   amidst of all changes that are going on around them.

   Information systems are typically used by distributed actors (both users and computers) who continually work with data.
   Every system contains a dataset, which represents the state of the system.
   Events that the system detects may cause the state to change.
   To keep our theory technology independent, we will assume that a dataset contains triples.
   This makes our theory valid for any kind of database that triples can represent,
   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.

@@

   For data migration, complications arise from accumulated data pollution in the old system.
   We distinguish data pollution in the following categories:
\begin{itemize}
   \item data pollution that cannot be captured in constraints on the dataset,
   such as a street address that has become obsolete because the system never registered the change of address.
   \item data pollution that violates a constraint in the schema.
   \item data pollution that violates a constraint that is not in the schema.
   \item data pollution that is a consequence of an erroneous constraint in the schema.
\end{itemize}
   The first category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
   In some cases, constraints on data can be formulated,
   so the system can help the user and simplify (or eliminate) the corresponding working procedure.
   So, such constraints may help to move this type of pollution to one of the other categories.
   The second category is about violations of constraints from the schema.
   Such violations do not occur in our theory because the old system was generated by Ampersand,
   so that system will keep all constraints from the schema satisfied at all times.
   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
   The third category, data pollution that is not captured by any of the constraints mentioned in the schema,
   can be dealt with by adding constraints in the new system.
   In that case, a migration engineer must ensure that the new system can start with data that satisfies all constraints in its schema.
   So some human intervention may be necessary in specific cases.
   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in the new system.

\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.35]{Migration.drawio.png}
   \end{center}
\caption{Data migration}
\label{fig:event flow}
\end{figure}
   This illustrates that automating data migrations has its limits.
   We cannot eliminate the need for interventions by a migration engineer entirely.
   If the semantics encoded in the new schema differs from the old schema,
   a migration engineer must make choices.
   If the interpretation of data changes,
   a migration engineer must remove obsolete constraints from the old system and introduce new constraints in the new system.
   The consequences of such changes may require changes in data as well.
   The migration engineer may even want to adapt existing data of which the constraints in the schema have not changed.
   All of this requires a third schema, which specifies the migration itself.
   This migration schema preserves data from the old system, it may introduce new data automatically (generated from the old dataset),
   and it may require users to introduce new data (which will take time). 
   It must contain the old schema, but it also contains the new schema and possibly some new data, all of which a migration engineer must specify.
   As the migration schema must distinguish between the old dataset and the new dataset,
   the migration must be based on the disjoint union of the two.
   So the idea of a migration schema is to compute the disjoint union of the old and the new schemas
   and to present it to the migration engineer in the form of source code,
   so the migration engineer can change everything she wants to suit the particulars of the data migration.

@@

   Atoms serve as data elements.
   Atoms are values without internal structure of interest, meant to represent data elements in a database.
   From a business perspective, atoms represent concrete items of the world,
   such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
   By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
   All atoms are taken from an infinite set called $\Atoms$.

@@

   The "normal" state of a system is that every rule in the schema is satisfied:
\begin{equation}
   \forall u\in\rules:\ \sat{u}{\dataset}\label{eqn:satisfaction}
\end{equation}
   However, the system continually registers events that insert or delete triples in its dataset,
   which may cause rules to be broken.
   This requires a reaction to restore invariance.
   Example: if a rule says that every request must be acknowledged, the receipt of a request (which is an event) will violate this rule.
   Sending a message that acknowledges this receipt (another event) will cause the rule to be satisfied again.
   During the time between the receipt of the request and sending the acknowledgment, the rule is (temporarily) not satisfied.

@@

\subsection{Example}
\label{old IS}
   Having defined an information system in mathematical terms, let us discuss a small example.
   For this purpose we use the language Ampersand
   because it makes the examples more appealing to read.
   Let us first define a dataset of just a handful of triples and three relations.
\begin{verbatim}
RELATION takes[Student*Course] =
[ ("Peter", "Management")
; ("Susan", "Business IT")
; ("John", "Business IT")
]
\end{verbatim}
   This declaration introduces a relation with name \verb#takes#,
   source concept \verb#Student#, and
   target concept \verb#Course#.
   The informal meaning of this relation is that it states which students are taking which courses.

@@

\subsection{Data migration}
   In a datamigration, we distinguish three moments:
\begin{enumerate}
   \item the moment of transition, i.e. the moment the old system is replaced (in production) by the new system;
   \item the decision point, i.e. the moment a business authority declares the migration successful;
   \item the point of no return, i.e. the moment that the migration cannot be undone other than by a backup-procedure,
         which is outside the scope of this paper.
\end{enumerate}
   The decision point must come after the moment of transition, to allow the business to verify the migration.
   The point of no return must come after the decision point, to allow for a no-go decision from the business.

   To migrate system $\infsys$ to $\infsys'$, we take the following steps:
\begin{enumerate}
   \item Let the old system $\infsys$ have state $\dataset$ and schema $\schema$.
         A developer defines a new system $\infsys'$, which contains a dataset $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.}.
         In Ampersand, this means that the developer produces an Ampersand script for the new system.
         The difference between the new system and the old system is the increment.
         The new system must contain the triples from $\dataset$ and $\dataset'$, to preserve the old data.
         If the developer wants to add some new data into the system on the moment of transition,
         she specifies that in $\dataset'$.
         That might be necessary to initialize specific new features.
         If $\schema$ equals $\schema'$, we call the migration trivial because a migration is not necessary.
   \item The generator generates a migration system $\migrsys$ based on $\schema$ and $\infsys'$,
         which gives the user the full functionality of the old system and the new system,
         i.e.\ $\schema_{\migrsys}=(\schema\sqcup\schema')\cup\id{migrationrules}$.
         The schema $\id{migrationrules}$ represents the enforce rules that migrate the existing data to the new dataset.
         TODO: $\id{migrationrules}$ bevat alleen tijdelijke datastructuur, indien nodig.
   \item A migration engineer changes $\id{migrationrules}$ to accommodate the non-standard requirements of the migration,
         yielding migration system $\migrsys'$.
         The migration engineer does not alter the old system $\infsys$ nor the specification new system $\infsys'$ that the developer has produced.
         As a result, the migration engineer can always restore the old system because it is part of the migration system.
   \item The production environment switches instantly from $\infsys$ to $\migrsys'$,
         while $\dataset_{\infsys}\subseteq\dataset_{\migrsys'}$.
         This means that the data of the old system remains intact.
         This event marks the moment of transition.
   \item The rules in the migration system that are not in $\schema'$ will remain in the system until they are satisfied,
         to enforce their execution.
         Automatic migration rules can be executed at the moment of transition.
         Migration rules that are enforced manually will be held alive until they are satisfied.
         All rules from $\schema$ (the rules from the old system) are removed on the moment of transition, because the are satisfied by property~\ref{eqn:satisfaction}.
   \item After a finite span of time, $\dataset_{\migrsys'}$ satisfies all its enforce rules.
         This implies that the generated rules plus the changes made by the migration engineer ensure the preservation of the meaning of the old data.
   \item Users may have to perform manual migration steps.
         These are specified as process rules in $\id{migrationrules}$.
         % Dit zijn "kurk" regels.
         The system ensures that the amount of work needed to perform these steps is finite.
   \item The patron of the system can decide whether to carry on with the migration or to roll back to the old system.
         She can do that incrementally by switching off the interfaces one-by-one.
   \item The system retains the data, the schema, and the functionality that is covered by $\schema'$, dropping everything else.
\end{enumerate}

@@

To implement the disjoint union, we need to relabel 
\begin{definition}[relabel concepts]
   \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
\end{definition}
\begin{definition}[relabel concepts in triples]
   \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
\end{definition}
\begin{definition}[relabel concepts in relations]
   \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
\end{definition}
\begin{definition}[relabel concepts in $\inst$]
   \[a\ \subst{C}{D}{\inst}\ A\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ a\ \inst\ D\ \text{\bf else}\ a\ \inst\ A\]
\end{definition}
\begin{definition}[relabel concepts in datasets]
   \[\begin{array}{rcl}
      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\inst}}
   \end{array}\]
\end{definition}
\begin{definition}[relabel concepts in rules]
   \[\sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}\]
\end{definition}
\begin{definition}[relabel concepts in schemas]
   \begin{eqnarray}
      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
      &=&\begin{array}[t]{l@{}l}
         \la&\{\subst{C}{D}{t}\mid t\in\triples\}\\
         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
         \end{array}
   \end{eqnarray}
\end{definition}

>>>>

\begin{abstract}
   To support incremental software development with a software generator bears the potential of decreasing the time to deploy changes.
   However, an increment that changes the database schema causes a nontrivial data migration problem.
   The problem is how to preserve the semantics of the data as much as possible and satisfy elementary business requirements at the same time.

   To generate an information system in a non-incremental manner is already
   provided for by the Ampersand project.
   This paper is about changing an already generated system in production.
   It develops a theory for deploying an incremental change,
   making Ampersand more useful for agile software development.
   We propose a migration method that features zero down-time,
   and features an undo after verification by the business.
   It also leaves room for the business to finish ongoing work
   after the moment the change has been deployed.
\end{abstract}

@@

   We believe in generating information systems%
\footnote{In the sequel, the word ``system'' refers to the phrase ``information system''. This simplifies the language a little. }
   to prevent human programming errors.
   For, a mistake not made can never propagate into production.
   We also believe that information systems should be released into production in small increments,
   to obtain shorter release cycles, more frequent releases, and thus a more agile software development process%
   \footnote{These effects can be measured in terms of DevOps metrics such as
   deployment frequency,
   reliability of deployments, and
   change failure rate~\cite{DevOps2021}.}.
   By installing upgrades automatically, frequently, in small increments, and without down-time,
   the users experience a system that evolves organically, without even noticing the updates.
   Thus, we believe our theory will help to automate data migrations correctly and reliably.
   The benefits work in two directions.
   Automating data migrations saves effort and errors,
   contributing to more frequent and smaller releases.
   Smaller and faster releases also mean a smaller difference between the old system and the new system.
   The smaller that difference, the simpler the migration can be and the more of it can be automated.

   In each increment, we distinguish an ``old system'' and a ``new system''.
   A difference of schemas complicates the migration of data between the old system and the new system.
   The purpose of incremental evolution poses specific requirements to data migration,
   which sets it apart from data migrations for other purposes.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers may exclude new functionality to avoid introducing new errors in an otherwise error-prone migration process.
   In our work, however, introducing new functionality and removing obsolete functionality is the very purpose of an increment.
   Data migration literature sometimes uses a different notion of increment.
   For example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014} define an increment as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   In our work, however, we want to get rid of the old system and not retain it.
   So, we need to look at a different approach.
   Examples like these have convinced us to define a new method for data migration that is specifically meant for systems in production that are
   evolving incrementally.

   The contribution of this paper is to derive a migration schema from two artifacts: an existing system
   (which has its own schema and dataset) and a new schema (which specifies the new system).
   The purpose of our theory is to develop tools.
   These tools must satisfy such practical requirements as
   zero down-time updates and ways to cope with data pollution.
   In practice, more frequent updates require that updates come with zero down-time.
   For our theory, zero down-time means that all human activity for the data migration can be done without taking the system out of production.
   We consider any delay caused by automatically switching from the old system to the new one negligible,
   assuming a modern deployment platform that does this with zero down-time.

@@

\section{Analysis}
   The purpose of an information system is to make data meaningful to its users.
   Users have their own tasks and responsibilities
   and may work from different locations and on different moments.
   This collective use by multiple users serves a purpose which we loosely call ``the business''.
   To preserve meaning, the business maintains semantic constraints on the data,
   amidst of all changes that are going on around them.

   Information systems are typically used by distributed actors (both users and computers) who continually work with data.
   As a consequence, the data in a system changes continually.
   Every system contains a dataset, which represents the state of the system.
   Events that the system detects may cause the state to change.
   To keep our theory technology independent, we will assume that a dataset contains triples.
   This makes our theory valid for any kind of database that triples can represent,
   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.

@@

   For data migration, complications arise from accumulated data pollution in the old system.
   We distinguish data pollution in the following categories:
\begin{itemize}
   \item data pollution that violates a constraint in the schema.
   \item data pollution that violates a constraint that is not in the schema.
   \item data pollution that cannot be captured in constraints on the dataset,
   such as a street address that has become obsolete because the system never registered the change of address.
   \item data pollution that is a consequence of an erroneous constraint in the schema.
\end{itemize}
   The first category is about violations of constraints from the schema.
   Such violations do not occur in our theory because the old system was generated by Ampersand,
   so the old system has kept all constraints from the schema satisfied.
   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
   The second category, data pollution that is not captured by any of the constraints mentioned in the schema,
   can be dealt with by adding constraints in the new system.
   In that case, a migration engineer must ensure that the new system can start with data that satisfies all constraints in its schema.
   So some human intervention may be necessary in specific cases.
   The third category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
   In some cases, constraints on data can be formulated,
   so the system can help the user and simplify (or eliminate) the corresponding working procedure.
   So, such constraints may help to move this type of pollution to one of the other categories.
   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in the new system.

\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.35]{Migration.drawio.png}
   \end{center}
\caption{Data migration}
\label{fig:event flow}
\end{figure}
   This illustrates that automating data migrations has its limits.
   We cannot eliminate the need for interventions by a migration engineer entirely.
   If the semantics encoded in the new schema differs from the old schema,
   a migration engineer sometimes has to make choices.
   If the interpretation of data changes,
   a migration engineer drops obsolete constraints from the old system and introduces new constraints in the new system.
   The consequences of such changes may require changes in data as well.
   The migration engineer may even want to adapt existing data of which the constraints in the schema have not changed.
   All of this requires a third schema, which specifies the migration itself.
   This migration schema preserves data from the old system, it may introduce new data automatically (generated from the old dataset),
   and it may require users to introduce new data (which will take time). 
   It must contain the old schema, but it also contains the new schema and possibly some new data, all of which a migration engineer must specify.
   As the migration schema must distinguish between the old dataset and the new dataset,
   the migration must be based on the disjoint union of the two.
   So the idea of a migration schema is to compute the disjoint union of the old and the new schemas
   and to present it to the migration engineer in the form of source code,
   so the migration engineer can change everything she wants to suit the particulars of the data migration.

@@

   Atoms serve as data elements.
   Atoms are values without internal structure of interest, meant to represent atomic data elements (e.g. dates, strings, numbers, etc.) in a database.
   From a business perspective, atoms represent concrete items of the world,
   such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
   By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
   All atoms are taken from an infinite set called $\Atoms$.

@@

   The system is {\em at rest} when every rule in the schema is satisfied:
\begin{equation}
   \forall u\in\rules:\ \sat{u}{\dataset}\label{eqn:satisfaction}
\end{equation}
   However, the system continually registers events that insert or delete triples in its dataset,
   which may cause rules to be broken.
   This requires a reaction to restore invariance.
   Example: if a rule says that every request must be acknowledged, the receipt of a request (which is an event) will violate this rule.
   Sending a message that acknowledges this receipt (another event) will cause the rule to be satisfied again.
   During the time between the receipt of the request and sending the acknowledgment, the rule is (temporarily) not satisfied.

@@

\subsection{Example}
\label{sct:Example old IS}
\label{old IS}
   Having defined an information system in mathematical terms, let us discuss a small example.
   For this purpose we use the language Ampersand
   because it makes the examples more appealing to read.
   Let us first define a dataset of just a handful of triples and three relations.
\begin{verbatim}
RELATION takes[Student*Course] =
[ ("Peter", "Management")
; ("Susan", "Business IT")
; ("John", "Business IT")
]
\end{verbatim}
   This declaration introduces a relation with name \verb#takes#,
   source concept \verb#Student#, and
   target concept \verb#Course#.
   The informal meaning of this relation is that it states which students are taking which courses.

@@

\subsection{Data migration}
   Sections~\ref{sct:Information Systems} and~\ref{sct:Example old IS} both show a complete information system.
   This section studies the transition from an old system to a new one, involving a data migration.

   In a datamigration, we distinguish three moments:
\begin{enumerate}
   \item the moment of transition, i.e. the moment the old system is replaced (in production) by the new system;
   \item the decision point, i.e. the moment a business authority declares the migration successful;
   \item the point of no return, i.e. the moment that the migration cannot be undone other than by a backup-procedure,
         which is outside the scope of this paper.
\end{enumerate}
   The decision point must come after the moment of transition, to allow the business to verify the migration.
   The point of no return must come after the decision point, to allow for a no-go decision from the business.

   To migrate system $\infsys$ to $\infsys'$, we take the following steps:
\begin{enumerate}
   \item Let the old system $\infsys$ have state $\dataset$ and schema $\schema$.
         A developer defines a new system $\infsys'$, which contains a dataset $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.}.
         In Ampersand, this means that the developer produces an Ampersand script for the new system.
         The {\em increment} is the difference between the new system and the old system.
         The new system must contain the triples from $\dataset$ and $\dataset'$, to preserve the old data.
         If the developer wants to add some new data into the system on the moment of transition,
         she specifies that in $\dataset'$.
         That might be necessary to initialize specific new features.
         If $\schema$ equals $\schema'$, we call the migration trivial because a migration is not necessary.
   \item The generator generates a migration system $\migrsys$ based on $\schema$ and $\infsys'$,
         which gives the user the full functionality of the old system and the new system,
         i.e.\ $\schema_{\migrsys}=(\schema\sqcup\schema')\cup\id{migrationrules}$.
         The schema $\id{migrationrules}$ represents the enforce rules that migrate the existing data to the new dataset.
         TODO: $\id{migrationrules}$ bevat alleen tijdelijke datastructuur, indien nodig.
   \item A migration engineer changes $\id{migrationrules}$ to accommodate the non-standard requirements of the migration,
         yielding migration system $\migrsys'$.
         The migration engineer does not alter the old system $\infsys$ nor the specification of the new system $\infsys'$ that the developer has produced.
         As a result, the migration engineer can always restore the old system because it is part of the migration system.
   \item The production environment switches instantly from $\infsys$ to $\migrsys'$,
         while $\dataset_{\infsys}\subseteq\dataset_{\migrsys'}$.
         This means that the data of the old system remains intact.
         This event marks the moment of transition.
   \item The rules in the migration system that are not in $\schema'$ will remain in the system until they are satisfied,
         to enforce their execution.
         Automatic migration rules can be executed at the moment of transition.
         Migration rules that are enforced manually will be held alive until they are satisfied.
         All rules from $\schema$ (the rules from the old system) are removed on the moment of transition, because the are satisfied by property~\ref{eqn:satisfaction}.
   \item After a finite span of time, $\dataset_{\migrsys'}$ satisfies all its enforce rules.
         This implies that the generated rules plus the changes made by the migration engineer ensure the preservation of the meaning of the old data.
   \item Users may have to perform manual migration steps.
         These are specified as process rules in $\id{migrationrules}$.
         % Dit zijn "kurk" regels.
         The system ensures that the amount of work needed to perform these steps is finite.
   \item The patron of the system can decide whether to carry on with the migration or to roll back to the old system.
         She can do that incrementally by switching off the interfaces one-by-one.
   \item The system retains the data, the schema, and the functionality that is covered by $\schema'$, dropping everything else.
\end{enumerate}

@@

To implement the disjoint union, we need to relabel 
\begin{definition}[relabel concepts]
   \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
\end{definition}
\begin{definition}[relabel concepts in triples]
   \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
\end{definition}
\begin{definition}[relabel concepts in relations]
   \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
\end{definition}
\begin{definition}[relabel concepts in $\inst$]
   \[a\ \subst{C}{D}{\inst}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\ \inst\ A\]
\end{definition}
\begin{definition}[relabel concepts in datasets]
   \[\begin{array}{rcl}
      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\inst}}
   \end{array}\]
\end{definition}
\begin{definition}[relabel concepts in rules]
   \[\sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}\]
\end{definition}
\begin{definition}[relabel concepts in schemas]
   \begin{eqnarray}
      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
      &=&\begin{array}[t]{l@{}l}
         \la&\{\subst{C}{D}{t}\mid t\in\triples\}\\
         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
         \end{array}
   \end{eqnarray}
\end{definition}
